Hypercoercions and a Framework for \\ Equivalence of Cast Calculi

### 2019-12-02 paper.tex
prove theorems in Agda
fix typos

\redrule{
  \highlight{\sOOreturn{v_3}{
    \kOOcaseIII{
        \vOOcast{v}{\cOOcast{\POOsum{T_1}{T_2}}{l}{\POOsum{T_3}{T_4}}}
      }{v_2}{\kappa}
  }}
}{
  \highlight{\sOOreturn{v_3'}{
      \kOOcaseIII{v}{v_2'}{\kappa}
  }}
}{\\&&
\highlight{\text{where}\;
v2' = \vOOcast{v_2}{\cOOcast{\POOfun{T_4}{T}}{l}{\POOfun{T_2}{T}}}}
\\&&
\highlight{\text{and}\;
v3' = \vOOcast{v_3}{\cOOcast{\POOfun{T_3}{T}}{l}{\POOfun{T_1}{T}}}}
}

  \redrule{
    \highlight{\sOOreturn{v}{
        \kOOcast{c}{\kappa}
      }}
  }{
\highlight{          
\begin{cases}
  \sOOreturn{v'}{\kappa} & \sidecond{applyCast(v,c) = \rOOsucc{v'}}
  \\
  \sOOhalt{\oOOblame{l}} & \sidecond{applyCast(v,c) = \rOOfail{l}}
\end{cases}}
  }{}
\redrule{
\sOOreturn{v}{\kOOmt}}{
\sOOhalt{observe(v}}{}
  \end{array}
  \]

@@

\begin{figure}
  Composition of hypercoercions \fbox{$c \fatsemi c = c$}
  \[
  \begin{array}{rclclr}
  c &\fatsemi& \hyperCoercionI{} &=& c\\
  \hyperCoercionI{} &\fatsemi& \hyperCoercionC{p_2}{m_2}{i_2} &=&
       \hyperCoercionC{p_2}{m_2}{i_2} \\
  \hyperCoercionC{p_1}{m_1}{\epsilon} &\fatsemi& \hyperCoercionC{\epsilon}{m_2}{i_2} &=&
       \hyperCoercionC{p_1}{m_1 \fatsemi m_2}{i_2} \\
  \hyperCoercionC{p_1}{m_1}{!} &\fatsemi& \hyperCoercionC{?^l}{m_2}{i_2} &=&
  \begin{cases}
    \hyperCoercionC{p_1}{m_1 \fatsemi m_2}{i_2} & \text{if } m_1 \smile m_2 \\
    \hyperCoercionC{p_1}{m_1}{\bot^l} & \text{otherwise}
  \end{cases} \\
  \hyperCoercionC{p_1}{m_1}{\bot^l} &\fatsemi& \hyperCoercionC{p_2}{m_2}{i_2} &=&
     \hyperCoercionC{p_1}{m_1}{\bot^l}
  \end{array}
  \]
  Composition of middles \fbox{$m \fatsemi m = m$}
  \[
  \begin{array}{rclclr}  
  \POOunit &\fatsemi& \POOunit &=& \POOunit \\
  c \to d &\fatsemi& c' \to d' &=& (c' \fatsemi c) \to (d \fatsemi d') \\
  c \times d &\fatsemi& c' \times d' &=& (c \fatsemi c') \times (d \fatsemi d') \\
  c + d &\fatsemi& c' + d' &=& (c \fatsemi c') + (d \fatsemi d')
  \end{array}
  \]
  Shallow consistency of middles \fbox{$m \smile m$}
  \[
  \POOunit \smile \POOunit \quad
  (c \to d) \smile (c' \to d') \quad
  (c \times d) \smile (c' \times d') \quad
  (c + d) \smile (c' + d')
  \]

@@

The $cast$ function is defined in Figure~\ref{fig:HC-UD-cast}.  Prior
presentations of this function do not use auxiliary functions, as we
do here. The reason that we introduction the auxiliary functions
$\mathit{castToDyn}$, $\mathit{castFromDyn}$, $\mathit{castToInj}$,
and $\mathit{castFromInj}$, is to ensure that each of them is
structurally recursive, which makes them straightforward to define in
Agda.

@@

\begin{lemma}[Weak Bisimulation between \effCEK{C} and \ineffCEKD{}]
  \label{thm:SD-bisim}
  Assume $C$ implements Lazy D Cast ADT
  and $C$ is a monoid
  and $cast(T,l,T)=id(T)$,
  $s_{1} \in \text{\effCEK{C}}$,
  $s_{2} \in \text{\ineffCEKD}$,
  $s_{1} \approx s_{2}$,
\begin{enumerate}
	\item If $\judgeSreduce{C}{s_1}{s_3}$ then
	$\judgeDreduceTrans{s_2}{s_4}$ and $s_3 \approx s_4$ for some $s_4$.
	\item If $\judgeDreduce{s_2}{s_4}$ then
	$\judgeSreduceTrans{C}{s_1}{s_3}$,
	$\judgeDreduceTrans{s_4}{s_5}$, and
	$s_3 \approx s_5$ for some $s_3$ and $s_5$.
	\item $s_1 = \sOOhalt{o}$ if and only if  $s_2 = \sOOhalt{o}$
\end{enumerate}
%\begin{enumerate}
%  \item $s_1 = \sOOhalt{o}$ if and only if  $s_2 = \sOOhalt{o}$
%%  \item If $s_2 = \sOOhalt{o}$ then $s_1 = \sOOhalt{o}$
%  \item If $\judgeSreduce{C}{s_1}{s_3}$ 
%    then $\judgeDreduce{s_2}{s_4}$ 
%    and  $s_3 \approx^{*} s_4$
%    for some $s_4$
%  \item If $\judgeDreduce{s_2}{s_4}$
%    then $\judgeSreduce{C}{s_1}{s_3}$ 
%    and  $s_3 \approx^{*} s_4$
%    for some $s_3$
%\end{enumerate}
%{\color{red}
%  Jeremy's attempt to reformulate in a more standard way that
%  remains equivalent:
%\begin{enumerate}
%\item If $\judgeSreduce{C}{s_1}{s_3}$ then
%  $\judgeDreducePlus{s_2}{s_4}$ and $s_3 \approx s_4$ for some $s_4$.
%\item If $\judgeDreduce{s_2}{s_4}$ then
%  $\judgeSreduce{C}{s_1}{s_3}$,
%  $\judgeDreduceTrans{s_4}{s_5}$, and
%   $s_3 \approx s_5$ for some $s_3$ and $s_5$.
%\item $s_1 = \sOOhalt{o}$ if and only if  $s_2 = \sOOhalt{o}$
%\end{enumerate}
%}
%Or diagrammatically, 
%  \[\begin{array}{clclc}
%  s_{1} & \longrightarrow_{\text{\effCEK{C}}} & s_1'\\
%  \rotatebox[origin=c]{90}{$\approx$} 
%  & & & \;\;\;\rotatebox[origin=c]{-20}{$\approx$} \\
%  s_{2} & 
%  \longrightarrow_{\text{\ineffCEKD}} & s_2' &
%  \longrightarrow^{*}_{\text{\ineffCEKD}} & s_2'' \\
%  \end{array}\]
\end{lemma}
\begin{proof}
  Part (1) follows from the definition of $s \approx s$.
  Part (2) and (3) are more interesting. 
  By the definition of $s \approx s$, it is straightforward to prove 
  $\judgeSreduce{C}{s_1}{s_3}$ if and only if $\judgeDreduce{s_2}{s_4}$.
  So we can prove two parts together by showing that 
  \[
  \text{if}\; s_{1} \approx s_{2}
  \;\text{and}\;\judgeSreduce{C}{s_1}{s_3}
  \;\text{and}\;\judgeDreduce{s_2}{s_4}
  \;\text{then}\; s_3 \approx^{*} s_4.
  \]
  To put it another way, our goal is to prove that 
  one reduction step in \effCEK{C} corresponds to
  one or more steps in \ineffCEKD.  
  Although there are many cases, the general idea is the same.
  Consider
  \[
  \sOOreturn{\hcvOOcons{v_{11}}{c_{11}}{v_{12}}{c_{12}}}{
    wrap(\kOOcar{\kappa})
  }
  \approx
  \sOOreturn{v_2}{\kOOcar{\kappa}}
  \]

@@

\subsubsection{\effCEK{C_1} and \effCEK{C_2} are equivalence if $C_1$ and $C_2$ are Lazy D}

@@

\begin{lemma}[$L$ is Correct wrt. \ineffCEKD]
  \label{lem:L-correct}
  If $\judgetype{\emptyset}{e}{T}$ and $o : T$, then
  \[
  \text{\evalEqv{\effCEK{L}}{\ineffCEKD}}
  \]
\end{lemma}
\begin{proof}
  Immediately from
  Correctness of \effCEK{C} (Lemma~\ref{thm:SD-equiv}) and 
  Properties of $L$ (Proposition~\ref{thm:L-LazyD-monoid}).
\end{proof}

>>>>

\redrule{
  \highlight{\sOOreturn{v_3}{
    \kOOcaseIII{
        \vOOcast{v}{\cOOcast{\POOsum{T_1}{T_2}}{l}{\POOsum{T_3}{T_4}}}
      }{v_2}{\kappa}
  }}
}{
  \highlight{\sOOreturn{v_3'}{
      \kOOcaseIII{v}{v_2'}{\kappa}
  }}
}{\\&&
\highlight{\text{where}\;
\kappa\;\text{is waiting for a value of type}\;T
}
\\&&
\highlight{\text{and}\;
	v2' = \vOOcast{v_2}{\cOOcast{\POOfun{T_4}{T}}{l}{\POOfun{T_2}{T}}}}
\\&&
\highlight{\text{and}\;
v3' = \vOOcast{v_3}{\cOOcast{\POOfun{T_3}{T}}{l}{\POOfun{T_1}{T}}}}
}

  \redrule{
    \highlight{\sOOreturn{v}{
        \kOOcast{c}{\kappa}
      }}
  }{
\highlight{          
\begin{cases}
  \sOOreturn{v'}{\kappa} & \sidecond{applyCast(v,c) = \rOOsucc{v'}}
  \\
  \sOOhalt{\oOOblame{l}} & \sidecond{applyCast(v,c) = \rOOfail{l}}
\end{cases}}
  }{}
\redrule{
\sOOreturn{v}{\kOOmt}}{
\sOOhalt{observe(v)}}{}
  \end{array}
  \]

@@

\begin{figure}
  Composition of hypercoercions \fbox{$c \fatsemi c = c$}
  \[
  \begin{array}{rclclr}
  c &\fatsemi& \hyperCoercionI{} &=& c\\
  \hyperCoercionI{} &\fatsemi& \hyperCoercionC{h_2}{m_2}{t_2} &=&
       \hyperCoercionC{h_2}{m_2}{t_2} \\
  \hyperCoercionC{h_1}{m_1}{\epsilon} &\fatsemi& 
  \hyperCoercionC{\epsilon}{m_2}{t_2} &=&
       \hyperCoercionC{h_1}{m_1 \fatsemi m_2}{t_2} \\
  \hyperCoercionC{h_1}{m_1}{!} &\fatsemi& \hyperCoercionC{?^l}{m_2}{t_2} &=&
  \begin{cases}
    \hyperCoercionC{h_1}{m_1 \fatsemi m_2}{t_2} & \text{if } m_1 \smile m_2 \\
    \hyperCoercionC{h_1}{m_1}{\bot^l} & \text{otherwise}
  \end{cases} \\
  \hyperCoercionC{h_1}{m_1}{\bot^l} &\fatsemi& \hyperCoercionC{h_2}{m_2}{t_2} 
  &=&
     \hyperCoercionC{h_1}{m_1}{\bot^l}
  \end{array}
  \]
  Composition of middles \fbox{$m \fatsemi m = m$}
  \[
  \begin{array}{rclclr}  
  \POOunit &\fatsemi& \POOunit &=& \POOunit \\
  c \to d &\fatsemi& c' \to d' &=& (c' \fatsemi c) \to (d \fatsemi d') \\
  c \times d &\fatsemi& c' \times d' &=& (c \fatsemi c') \times (d \fatsemi d') \\
  c + d &\fatsemi& c' + d' &=& (c \fatsemi c') + (d \fatsemi d')
  \end{array}
  \]
  Shallow consistency of middles \fbox{$m \smile m$}
  \[
  \POOunit \smile \POOunit \quad
  (c \to d) \smile (c' \to d') \quad
  (c \times d) \smile (c' \times d') \quad
  (c + d) \smile (c' + d')
  \]

@@

The $cast$ function is defined in Figure~\ref{fig:HC-UD-cast}.  Prior
presentations of this function do not use auxiliary functions, as we
do here. The reason that we introduce the auxiliary functions
$\mathit{castToDyn}$, $\mathit{castFromDyn}$, $\mathit{castToInj}$,
and $\mathit{castFromInj}$, is to ensure that each of them is
structurally recursive, which makes them straightforward to define in
Agda.

@@

\begin{lemma}[Weak Bisimulation between \effCEK{C} and \ineffCEKD{}]
  \label{thm:SD-bisim}
  Assume $C$ implements Lazy D Cast ADT
  and $C$ is a monoid
  and $cast(T,l,T)=id(T)$,
  $s_{1} \in \text{\effCEK{C}}$,
  $s_{2} \in \text{\ineffCEKD}$,
  $s_{1} \approx s_{2}$,
\begin{enumerate}
	\item If $\judgeSreduce{C}{s_1}{s_3}$ then
	$\judgeDreduceTrans{s_2}{s_4}$ and $s_3 \approx s_4$ for some $s_4$.
	\item If $\judgeDreduce{s_2}{s_4}$ then
	$\judgeSreduceTrans{C}{s_1}{s_3}$,
	$\judgeDreduceTrans{s_4}{s_5}$, and
	$s_3 \approx s_5$ for some $s_3$ and $s_5$.
	\item $s_1 = \sOOhalt{o}$ if and only if  $s_2 = \sOOhalt{o}$
\end{enumerate}
%\begin{enumerate}
%  \item $s_1 = \sOOhalt{o}$ if and only if  $s_2 = \sOOhalt{o}$
%%  \item If $s_2 = \sOOhalt{o}$ then $s_1 = \sOOhalt{o}$
%  \item If $\judgeSreduce{C}{s_1}{s_3}$ 
%    then $\judgeDreduce{s_2}{s_4}$ 
%    and  $s_3 \approx^{*} s_4$
%    for some $s_4$
%  \item If $\judgeDreduce{s_2}{s_4}$
%    then $\judgeSreduce{C}{s_1}{s_3}$ 
%    and  $s_3 \approx^{*} s_4$
%    for some $s_3$
%\end{enumerate}
%{\color{red}
%  Jeremy's attempt to reformulate in a more standard way that
%  remains equivalent:
%\begin{enumerate}
%\item If $\judgeSreduce{C}{s_1}{s_3}$ then
%  $\judgeDreducePlus{s_2}{s_4}$ and $s_3 \approx s_4$ for some $s_4$.
%\item If $\judgeDreduce{s_2}{s_4}$ then
%  $\judgeSreduce{C}{s_1}{s_3}$,
%  $\judgeDreduceTrans{s_4}{s_5}$, and
%   $s_3 \approx s_5$ for some $s_3$ and $s_5$.
%\item $s_1 = \sOOhalt{o}$ if and only if  $s_2 = \sOOhalt{o}$
%\end{enumerate}
%}
%Or diagrammatically, 
%  \[\begin{array}{clclc}
%  s_{1} & \longrightarrow_{\text{\effCEK{C}}} & s_1'\\
%  \rotatebox[origin=c]{90}{$\approx$} 
%  & & & \;\;\;\rotatebox[origin=c]{-20}{$\approx$} \\
%  s_{2} & 
%  \longrightarrow_{\text{\ineffCEKD}} & s_2' &
%  \longrightarrow^{*}_{\text{\ineffCEKD}} & s_2'' \\
%  \end{array}\]
\end{lemma}
\begin{proof}
  Part (3) follows from the definition of $s \approx s$.
  Part (1) and (2) are more interesting. 
  By the definition of $s \approx s$, it is straightforward to prove 
  $\judgeSreduce{C}{s_1}{s_3}$ if and only if $\judgeDreduce{s_2}{s_4}$.
  So we can prove two parts together by showing that if $s_{1} \approx s_{2}$
  and $\judgeDreduce{s_2}{s_4}$ then
  \[
%  \text{if}\; 
%  \;\text{and}\;\judgeSreduce{C}{s_1}{s_3}
%  \;\text{and}\;\judgeDreduce{s_2}{s_4}
  \;\text{there exists}\; s'
  \;\text{such that}\; \judgeDreduceTrans{s_4}{s'}
  \;\text{and}\; s_3 \approx s'.
  \]
  To put it another way, our goal is to prove that 
  one reduction step in \effCEK{C} corresponds to
  one or more steps in \ineffCEKD.  
  Although there are many cases, the general idea is the same.
  Consider
  \[
  \sOOreturn{\hcvOOcons{v_{11}}{c_{11}}{v_{12}}{c_{12}}}{
    wrap(\kOOcar{\kappa})
  }
  \approx
  \sOOreturn{v_2}{\kOOcar{\kappa}}
  \]

@@

\subsubsection{\effCEK{C_1} and \effCEK{C_2} are equivalent if $C_1$ and $C_2$ 
are Lazy D}

@@

\begin{lemma}[$L$ is Correct wrt. \ineffCEKD]
  \label{lem:L-correct}
  If $\judgetype{\emptyset}{e}{T}$ and $o : T$, then
  \[
  \text{\evalEqv{\effCEK{L}}{\ineffCEKD}}
  \]
\end{lemma}
\begin{proof}
  Immediately from
  Correctness of \effCEK{C} (Corollary~\ref{thm:SD-equiv}) and 
  Properties of $L$ (Proposition~\ref{thm:L-LazyD-monoid}).
\end{proof}
